---
title: "BIOF 339: Practical R"
output:
  pdf_document: default
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F, warning=F, message = F)
library(tidyverse)
library(lubridate)
library(knitr)
library(pander)

startDate <- lubridate::ymd('20190911')
classDate <- function(wk, startDate ){
  today <- startDate + 7*(wk-1)
  return(format(today, "%B %d, %Y"))
}
```

## Instructor 
Abhijit Dasgupta  
Contact: via Slack or email ([adasgupta+biof339@araastat.com](mailto:adasgupta+biof339@araastat.com)) 

## Course description

The goal of this course is to introduce R as an analysis platform and tool for data science rather than a programming language.
Throughout the course, emphasis will be placed on example-driven
learning. Topics to be covered include: installation of R and R
packages; command line R; R data types; loading data in R; manipulating
data; exploring data through visualization; statistical tests;
correcting for multiple comparisons; building models;  generating
publication-quality graphics; creating reports using RMarkdown. No prior programming experience is
required.

# Learning Objectives

- Run R and RStudio, making use of inherent R features
- Find and make use of the extensive packages (R add-ons) available for analyzing biological and other forms of data
- Load, manipulate, and combine data to make it amenable to further analyses
- Visualize data with extensive graphics capabilities of R (including ggplot)
- Use R to run statistical models and hypothesis tests and report results conforming to standards expected in scientific journals
- Write reports using the powerful `rmarkdown` package and its derivatives

## Required computers and software

You are required to bring to each class a personal laptop running Windows, Mac OS X or Linux. You 
are also required to install the software [R](http://cran.r-project.org){target=_blank} and
the integrated development environment [RStudio](https://www.rstudio.com){target=_blank}. Instructions for 
installing these are available on the Resources page (see links above).


## Outline of the class

```{r outline, echo = FALSE}
dts <- classDate(1:14, startDate)
topics <- c(
  'Introduction to R, RStudio and RMarkdown',
  'Data Structures in R (classes 5:30-7, 7-8:30)',
  'R packages, data import/export, munging',
  'Towards analytic data: Data Munging, continued',
  'Data exploration through visualization',
  'More data visualization and RMarkdown',
  'Statistical analyses: Table 1, estimation and confidence intervals, and more ggplot',
  'Statistical analyses: Classical hypothesis testing and computational inference',
  'Statistical learning: Regression models',
  "More data munging with `purrr`: grouping, mapping and functional programming",
  "Basic bioinformatics: Bioconductor and friends",
  "No class (Thanksgiving)",
  'Statistical learning: Cluster analysis and pattern recognition',
  "Project presentations")
D <- tibble(Date = dts, Topic=topics)
knitr::kable(D)

```

## Books and learning materials

There are no required books for this class. However, we will extensively refer to
a few books available freely online and will serve as reading material and ongoing 
reference material for this course. 

1. _R for Data Science_ [R4DS] by Hadley Wickham and Garrett Grolemund (available [online](https://r4ds.had.co.nz/){target=_blank})

### Communication

This class will communicate primarily via [Slack](http://www.slack.com){target=_blank}. Please join the BIOF339 Slack channel using [this link](https://join.slack.com/t/biof339/shared_invite/enQtNzU1Njk5NDc2MDMyLWU5MjkwNzNlMmRlYTAwNDg2MGZjYWQ1YzEzMDUxMzc0MzE1NjZkMDFlY2Y0NTIyMGI2Y2VkODBiYmY0MjBiMzE).

You will see two channels named *wed5-7_2019* and *wed7-9_2019*. Please join the channel corresponding to your section. I will be using Slack for broadcasting messages, answering questions and the like. If you have a question, you can directly message me on Slack if you like. Expect an answer within 24 hours.

## Grades

Grades will be based on the following requirements:

1. Homeworks, available Friday after class, due by 11:59PM the following Tuesday.  (50%)
    - No late homeworks, since solutions will be available Wednesday mornings
    - I'll score the top 10 homeworks for grade
1. Final project: A RMarkdown report/presentation demonstrating an end-to-end data analysis in R using your own data, from data ingestion to munging to analyses and graphics, with a brief introduction and conclusion (20%)
1. Class participation (20%)
1. Completion and submission of class exercises (10%, marked for completion)
    - These will need to be in basic RMarkdown, showing the problem and the solution. You can add a section for questions here that I can address in the following class or online. These will have to be submitted before you leave the classroom

## Academic policy regarding plagiarism

The FAES Graduate School at NIH prides itself on providing quality
educational experiences and upholds the highest level of honesty,
integrity, and mutual respect. It is our policy that cheating,
fabrication or plagiarism by students is\
not acceptable in any form. If a student is found to be in violation of
any, or all of the below, his/her credits will be forfeited, and he/she
will not be allowed to enroll in future courses or education programs
administered by FAES.

• Cheating is defined as an attempt to give or obtain inappropriate/
unauthorized assistance during any academic exercise, such as during
examination, homework assignment, class presentation.\
• Fabrication is defined as the falsification of data, information or
citations in any academic materials.\
• Plagiarism is defined as using the ideas, methods, or written words of
another, without proper acknowledgment and with the intention that they
be taken as the work of the deceiver. These include, but are not limited
to, the use of published articles, paraphrasing, copying someone else’s
homework and turning it in as one’s own and failing to reference
footnotes. Procuring information from online sources without proper
attribution also constitutes plagiarism.

See [this link](https://faes.org/sites/default/files/Student_Handbook_2016-17.pdfhttps://faes.org/sites/default/files/Student_Handbook_2016-17.pdf){target=_blank} for FAES policy on academic plagiarism.


