---
title: "Homework 1 Document"
author: "Dr. Dasgupta"
output: html_document
---

The data `airquality` comprises daily air quality measurements in New York, 
May to September, 1973. 

The table below shows the average monthly temperature in New York during this period

```{r}
library(tidyverse)
library(knitr)
avg_temp_by_month <- airquality %>% 
  group_by(Month) %>% 
  summarize(avgTemp = mean(Temp, na.rm=T))
kable(avg_temp_by_month)
```

The pattern of monthly temperatures is plotted below:

```{r}
ggplot(avg_temp_by_month, aes(x = Month, y = avgTemp)) +
  geom_point() + 
  geom_line(color = 'blue') + 
  labs(y = 'Average Temperature (F)')
```

Wind may have an effect on the amount of Ozone in the air. We plot the 
relationship between recorded wind speed and ozone levels below:

```{r}
ggplot(airquality, aes(x = Wind, y = Ozone)) + 
  geom_point() + 
  geom_smooth(color = 'blue', se = FALSE)
```

We find the maximum recorded level during this period was 
`r max(airquality$Ozone, na.rm = T)` ppb. 
