---
title: "Spotify Wrapped (Apple Music Style)"
author: "Hayley Sowards"
output: ioslides_presentation
---

```{r setup, include = F}
knitr::opts_chunk$set(message = F, warning = F)
library(readr)
library(tidyverse)
library(knitr)
library(rjson)

df <- read_csv("data/Apple Music Play Activity.csv")
```

## The Data

- [Apple has a way to request your listening data ](https://privacy.apple.com/) 
- Comes with all of these variables: `r colnames(df)`

## Spotify's Wrapped Includes:

- Top 5 songs
- Top 5 artists 
- Total time listened (minutes)
- Top genre
- Decade wrapped (minutes listened, top song, and top artist)

## Clean Data

```{r, echo = F}
music <- df %>% 
  mutate(artist = `Artist Name`, song = `Content Name`, genre = `Genre`, media_duration = `Media Duration In Milliseconds`, play_duration = `Play Duration Milliseconds`, time_start = `Event Start Timestamp`) %>% #renaming all of these for ease of analysis down the line
  select(artist, song, genre, media_duration, play_duration, time_start) %>% #pulling only needed vars
  mutate_if(is.character, as.factor) %>% #you know, the usual
  mutate(media_duration = media_duration/1000, play_duration = play_duration/1000) %>% #conversion of milliseconds to seconds
  filter(!is.na(artist), !is.na(song)) #remove songs without a title or artist

music2 <- music %>% #cleaning dates and finalizing dataset, using this dataset for decades analysis
  separate(time_start, into = c("year", "month", "datetime"), sep = "-") %>% 
  separate(datetime, into = c("date", "time"), sep = " ") %>% 
  mutate_if(is.character, as.numeric) %>% 
  filter(artist != "GentleWhispering")  #this is an artist I have in my library that posts videos as audio, I'm taking her out

head(music2)

music2019 <- music2 %>% #used for 2019 analyses
  filter(year == 2019)
```

- `dplyr::select` out relevant variables
- `dplyr::separate` out date variable
- `dplyr::filter` out songs from this year

## Top Genre

```{r}
genre <- music2019 %>% 
  group_by(genre) %>% 
  summarise(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  #using %/% when I want the minutes displayed to avoid decimals
  arrange(desc(Total.Play))
kable(genre[1,1:2])
```

## Minutes Listened

```{r}
time2019 <- sum(music2019$play_duration, na.rm = T)%/%60
time2019 <- as.character(time2019)
```

`r time2019` minutes

## Top 5 Artists

```{r}
artistbytime <- music2019 %>% 
  group_by(artist) %>% 
  summarise(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  arrange(desc(Total.Play))
```
```{r, echo = F}
kable(artistbytime[1:5,1:2])
```

## Top 5 Songs

```{r}
songbytime <- music2019 %>% 
  group_by(song, artist) %>% 
  summarise(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  arrange(desc(Total.Play))
```
```{r, echo = F}
kable(songbytime[1:5,1:3])
```

## Decade Wrapped: Artist of the Decade

```{r}
artistsdecade <- music2 %>%
  group_by(artist) %>% 
  summarise(Total.Play = (sum(play_duration, na.rm = T)/60)) %>% 
  arrange(desc(Total.Play))
```

```{r, echo = F}
kable(artistsdecade[1,])
```


- `r artistsdecade[1,2]%/%60` hours and `r artistsdecade[1,2]/1440` days

## Decade Wrapped: Song of the Decade

```{r}
songsdecade <- music2 %>%
  group_by(song, artist) %>%
  summarise(Total.Play = (sum(play_duration, na.rm = T)/60)) %>% 
  arrange(desc(Total.Play))
```

```{r, echo = F}
kable(songsdecade[1,])
```

## Decade Wrapped: Time Each Year

```{r}
musicyearly <- music2 %>% 
  group_by(year) %>% 
  summarize(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  filter(!is.na(year)) %>% 
  arrange(desc(year))
```
```{r, echo = F}
kable(musicyearly)
```

## Play Duration Each Year

```{r}
p_yearly <- ggplot(musicyearly, aes(x = year, y = Total.Play)) +
  geom_point() +
  geom_segment(aes(x = year, xend = year, y = 0, yend = Total.Play)) +
  labs(x = "Year", y = "Play Duration in Minutes") +
  theme_minimal()
```

## Play Duration Each Year

```{r, echo = F}
p_yearly
```

## Decade Wrapped: Top Artist Each Year
```{r}
artistyearly <- music2 %>% 
  group_by(year, artist) %>% 
  summarize(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  filter(!is.na(year)) %>% 
  top_n(1, Total.Play) %>% 
  arrange(desc(year))
```
```{r, echo = F}
kable(artistyearly)
```

## Decade Wrapped: Top Song Each Year
```{r}
songyearly <- music2 %>% 
  group_by(year, song, artist) %>% 
  summarize(Total.Play = (sum(play_duration, na.rm = T)%/%60)) %>% 
  filter(!is.na(year)) %>% ungroup() %>% group_by(year) %>% 
  top_n(1, Total.Play) %>% 
  arrange(desc(year))
```
```{r, echo = F}
kable(songyearly)
```

##

```{r}
songs <- music2 %>%
  group_by(song, artist, genre) %>% 
  filter(!is.na(song), !is.na(genre), !is.na((artist))) %>% 
  summarize(Total.Play = (sum(play_duration, na.rm = T)/3600)) %>%
  arrange(desc(Total.Play))

id <- 1:20

top20 <- songs[1:20,] %>% 
  arrange(song) %>% 
  add_column(id)

# Get the name and the y position of each label
label_top20 <- top20
# calculate the ANGLE of the labels
number_of_bar <- nrow(label_top20)
angle <-  90 - 360 * (label_top20$id-0.5) /number_of_bar
# calculate the alignment of labels: right or left
# flip angle BY to make them readable
label_top20$hjust<-ifelse( angle < -90, 1, 0)
label_top20$angle<-ifelse(angle < -90, angle+180, angle)
```

##

```{r}
p_songs <- ggplot(top20, aes(x=as.factor(song), y=Total.Play)) +
  geom_col(aes(fill = artist)) +
  ylim(0,50) + #first number changes inner circle (0, no inner circle, negative = larger), second number changes outer circle
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-2,4), "cm")) +
  coord_polar(start = 0) +
  geom_text(data=label_top20, aes(x=id, y=Total.Play+5, label=song, hjust=hjust), 
            color="black", fontface="bold",alpha=0.6, size=3.5, angle= label_top20$angle, inherit.aes = FALSE )
```

## Top 20 Songs of the Decade (but Circular)

```{r, echo = F}
p_songs
```


## Hours of Music Everyday (but in a Manhattan Plot)

```{r}
p_manhattan <- music2 %>% 
  mutate(year = as.factor(year)) %>% 
  group_by(year, month, date) %>% 
  arrange(year, month, date) %>%
  mutate(chron = paste(year,month, date, sep = "-")) %>% 
  mutate(Daily.Play = sum(play_duration)) %>%
  filter(!is.na(year)) %>% 
  ggplot(aes(x = chron, y = (Daily.Play/60), group = year, color = year)) +
  geom_point() +
  labs(x = "Date", y = "Duration in Minutes") +
  theme(axis.text.x = element_blank())
```


## Hours of Music Everyday (but in a Manhattan Plot)
```{r,echo=F}
p_manhattan
```
