---
title: "Homework 4: Data munging"
author: "Solutions"
date: ""
output: 
  html_document:
    code_folding: show
---

```{r setup, include=F}
knitr::opts_chunk$set(warning=F, message=F)
```

## Homework

The dataset [here](data/clinical_data_breast_cancer_hw.csv) provides clinical and
pathological data for 105 breast cancer patients. Create a fresh RMarkdown file and answer the following questions:

<OL>
<li>Read this data into R, calling it `brca_data`. 

```{r}
library(rio)
brca_data <- import('data/clinical_data_breast_cancer_hw.csv') # You may have a different file name 
```

<li>This data is in tidy format, but needs cleaning. 
<ol type='a'>
<li>Evaluate whether all the variables are of the correct type. If not, evaluate why, and correct the problem.

```{r, message=F}
library(tidyverse)
str(brca_data)
brca_data <- # Over-write original data
  brca_data %>% 
  mutate(Age.at.Initial.Pathologic.Diagnosis = as.numeric(Age.at.Initial.Pathologic.Diagnosis))
```
    
<li>For the molecular results for ER, PR and HER2, any entry that is not Positive or Negative should be changed to a missing value.

```{r}
brca_data <- 
  brca_data %>% 
  mutate(ER.Status = ifelse(ER.Status != "Positive" & ER.Status != 'Negative', NA, ER.Status),
         PR.Status = ifelse(PR.Status != 'Positive' & PR.Status != 'Negative', NA, PR.Status),
         HER2.Final.Status = ifelse(HER2.Final.Status != 'Positive' & HER2.Final.Status != 'Negative',NA, HER2.Final.Status))
```
</ol>
<li>Create a new dataset that contains patient ID, ER, PR and HER2 status. You need to use one of the functions `starts_with`, `ends_with` or `contains` within the `select` function. 
```{r}
brca_molecular <- brca_data %>% 
  select(Complete.TCGA.ID, ends_with('Status'))
```

<li> Creating new variables (what `dplyr` function will you use for all of these?):
<ol type='a'>
<li>Create a variable giving the TNM status of each patient. The T, N and M statuses are given separately. I want a single variable encoded as, for example, "T2N0M0". [Hint: The function `paste` is your friend]

```{r}
brca_data <- brca_data %>% 
  mutate(TNM = paste(Tumor, Node, Metastasis, sep=''))
```

A (perhaps nicer) solution is 

```{r}
brca_data <- brca_data %>% 
  unite(TNM, Tumor, Node, Metastasis, sep='', remove = F)
```

<li>Create a variable that divides the patients into 3 categories based on molecular profiles [Hint: The function `case_when` might help]:
<ul>
<li> Luminal (ER positive and/or PR positive)
<li> HER2 (HER2 positive)
<li> Basal-like (ER, PR and HER2 negative)
</ul>

```{r}
brca_data <- brca_data %>% 
  mutate(brca_cat = case_when(
    ER.Status=='Positive' | PR.Status == 'Positive' ~ 'Luminal',
    HER2.Final.Status=='Positive' ~ 'HER2',
    ER.Status=='Negative' & PR.Status=='Negative' & HER2.Final.Status=='Negative' ~ 'Basal-like'
  ))
# This makes tumors with (either ER+ or PR+) and HER2+ to be Luminal. Why?
```
        
<li>Create a variable that equals the time of death if they are dead (see the variable `Vital.Status`) or the time of last contact if they are alive. This is a common computation for survival analysis studies, and is called the overall survival time. [Hint: the function `ifelse` might be useful]

```{r}
brca_data <- brca_data %>% 
  mutate(OS.Survival = ifelse(Vital.Status=='DECEASED',
                              Days.to.date.of.Death,
                              Days.to.Date.of.Last.Contact))
```
</ol>
<li>Let's compute some summaries 
<ol type='a'>
<li> Find the number of patients with each breast cancer type
<li> Find the average age of patients with each breast cancer type
<li> Find the proportion of patients who are dead for each breast cancer type
<li> Find the proportion of patients with each TNM status
</ol>
</ol>

> c and d require creating a new variable to compute the proportion after determining the counts using either `count` or a group_by/summarize pipe using `n()`.

```{r}
brca_data %>% count(brca_cat)
brca_data %>% 
  group_by(brca_cat) %>% 
  summarize(avgAge = mean(Age.at.Initial.Pathologic.Diagnosis, na.rm=T))
brca_data %>% 
  group_by(brca_cat) %>% 
  summarize(prop_dead = mean(Vital.Status=='DECEASED'))
brca_data %>% 
  count(TNM) %>% 
  mutate(prop = n/sum(n)) %>% 
  select(-n) %>% 
  arrange(desc(prop)) # Optional
```

**[Extra credit:]** What proportion of the Luminal patients are of tumor stage T1 or T2?

```{r}
brca_data %>% 
  filter(brca_cat=='Luminal') %>% 
  summarize(early_stage = mean(Tumor == 'T1' | Tumor == 'T2', na.rm=T))
```


Save the cleaned breast cancer dataset as `brca_cleaned`. You can save this to your computer using `saveRDS(brca_cleaned, file="<a filename of your choice>.rds")`. We'll be using this dataset again whe we do plots and modeling.

```{r}
brca_cleaned <- brca_data # Since this is after we've done everything above
saveRDS(brca_cleaned, file = 'data/brca.rds')
```


    
