---
title: "Describing and summarizing data"
author: Abhijit Dasgupta
date: BIOF 339

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '#> ')
library(tidyverse)
```

layout: true

<div class="my-header">
<span>BIOF339, BIOF 339</span></div>

---

## Where we've been

1. Understand what tidy data is
1. Manipulate data to make it tidy (tidyr, dplyr)
1. Transform particular variables
1. Write basic functions
1. High-throughput analyses
    - Lists of data sets
    - `map` to apply similar processes to each data set
    - for-loops to repeat same recipe on multiple data sets or objects

---

## Where we're going

1. Creating data summaries
1. Basic statistical comparisons between groups
1. Creating tables 
    -  Table 1
    -  Tables for analytic results

--

The basic assumption we'll make is that we will start with a tidy data set.


---
class: middle, center

# Statistical summaries

---

## Univariate summaries

**Single summaries**  
.pull-left[
- Mean (`mean`)
- Variance(`var`)
- Standard deviation (`sd`)
- Count (`nrow` or `dplyr::n` or `dplyr::n_distinct`)
]
.pull-right[
- Median ('median')
- Inter-quartile range (`IQR`)
- Mean absolute deviation (`mad`)
- Minimum (`min`) and Maximum (`max`)
]

--

**Multiple summaries**  
- Quantiles (`quantile`)
- Range (`range`)

---
class: middle, center

# Summarizing the breast cancer expression dataset

---

## Mean

.pull-left[
```{r s1, eval = F, echo = T}
brca <- rio::import('data/BreastCancer_Expression.csv')
brca %>% 
  summarize_at(across(starts_with('NP'), 
               mean, na.rm=T)) #<<
```
]
.pull-right[
```{r 08-Summaries-1, eval=T, echo = F, ref.label="s1"}
```
]

---

## Median

.pull-left[
```{r s2, eval = F, echo = T}
brca %>% 
  summarize_at(vars(starts_with('NP')), 
               median, na.rm=T) #<<
```
]
.pull-right[
```{r 08-Summaries-2, eval=T, echo = F, ref.label="s2"}
```
]

---

## Standard deviation

.pull-left[
```{r s3, eval = F, echo = T}
brca %>% 
  summarize_at(vars(starts_with('NP')),
               sd, na.rm=T) #<<
```
]
.pull-right[
```{r 08-Summaries-3, eval=T, echo = F, ref.label="s3"}
```
]

---

## Multiple summaries together

.pull-left[
```{r s4, eval = F, echo = T}
brca %>% 
  summarize_at(vars(starts_with('NP')),
               c(mean,
               median,
               sd), na.rm=T)
```
]
.pull-right[
```{r 08-Summaries-4, eval=T, echo = F, ref.label="s4"}
```
]

---

## Multiple summaries together

.pull-left[
```{r s5, eval = F, echo = T}
brca %>% 
  summarize_at(-1, # got tired of typing
               c('Mean'=mean,
                 'Median' = median,
                 'SD'=sd), na.rm=T) 
```
]
.pull-right[
```{r 08-Summaries-5, eval=T, echo = F, ref.label="s5"}
```
]

---

## Multiple summaries together

.pull-left[
```{r s7, eval = F, echo = T}
brca %>% 
  summarize_at(-1, 
               c('Mean' = mean,
                 'Median' = median,
                 'SD' = sd), na.rm=T) %>% 
  tidyr::gather(variable, value) %>% #<<
  separate(variable, #<<
           c("Type",'ID','Statistic'), sep='_') %>% #<<
  spread(Statistic, value) %>% #<<
  unite(ID, c('Type','ID'), sep='_') #<<
```

The highlighted part is to format the output
]
.pull-right[
```{r 08-Summaries-6, eval=T, echo = F, ref.label="s7"}
```
]

---

## Data set summary

There is a function `summary` that will give you summaries of all the variables. It's nice for looking at the data, but the output format isn't very good for further manipulation

.pull-left[
```{r s8, eval = F, echo = T}
summary(brca[,-1])
```
]
.pull-right[
```{r 08-Summaries-7, eval=T, echo = F, ref.label="s8"}
```
]

---
class: middle, center

# Maybe an easier way?

---

## The `tableone` package

The `tableone` package is meant to create, you guessed it, Table 1. 

It is quite a convenient package for most purposes and saves gobs of time

---

## The `tableone` package

.pull-left[
```{r t1, eval = F, echo = T}
library(tableone)
tab1 <- CreateTableOne(data=brca[,-1])
tab1
```
]
.pull-right[
```{r 08-Summaries-8, eval=T, echo = F, ref.label="t1"}
```
]

---

## The `tableone` package

.pull-left[
```{r t2, eval = F, echo = T}
library(tableone)
tab1 <- CreateTableOne(data = brca[-1])
print(tab1, nonnormal = names(brca)[-1])
```

You have to give the variable names of those you think are non-normally distributed and need to be summarized by the median
]
.pull-right[
```{r 08-Summaries-9, eval=T, echo = F, ref.label="t2"}
```
]

---

## The `tableone` package

.pull-left[
```{r t3, eval = F, echo = T}
library(tableone)
tab1 <- CreateTableOne(data = brca[-1])
kableone(print(tab1, nonnormal = names(brca)[-1]),
         format='html')
```
]
.pull-right[
```{r 08-Summaries-10, eval=T, echo = F, ref.label="t3"}
```
]

---
class: middle, center

# Mixed data

---

Let's first put the expression and clinical data together

```{r 08-Summaries-11}
library(rio)
brca1 <- import('data/clinical_data_breast_cancer_hw.csv')
brca2 <- import('data/BreastCancer_Expression.csv')
brca <- left_join(brca1, brca2, by=c('Complete.TCGA.ID' = 'TCGA_ID')) %>% 
  mutate(Age.at.Initial.Pathologic.Diagnosis =
           as.numeric(Age.at.Initial.Pathologic.Diagnosis)) %>% 
  mutate(ER.Status = ifelse(ER.Status %in% c('Positive','Negative'),
                            ER.Status, NA))

summary(brca)
```

---

Let's first put the expression and clinical data together

```{r 08-Summaries-12}
library(rio)
brca1 <- import('data/clinical_data_breast_cancer_hw.csv')
brca2 <- import('data/BreastCancer_Expression.csv')
brca <- left_join(brca1, brca2, by=c('Complete.TCGA.ID' = 'TCGA_ID')) %>% 
  mutate(Age.at.Initial.Pathologic.Diagnosis =
           as.numeric(Age.at.Initial.Pathologic.Diagnosis)) %>% 
  mutate(ER.Status = ifelse(ER.Status %in% c('Positive','Negative'),
                            ER.Status, NA),
         HER2.Final.Status = ifelse(HER2.Final.Status=='Equivocal', 
                                    NA, HER2.Final.Status)) %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate(Complete.TCGA.ID = as.character(Complete.TCGA.ID)) #<<

str(brca)
```

---

Identify which variables are categorical (factors) and which are continuous (numeric)

```{r 08-Summaries-13}
catvars <- brca %>% select_if(is.factor) %>% names()
ctsvars <- brca %>% select_if(is.numeric) %>% names()
```

---

.pull-left[
```{r tab1, eval = F, echo = T}
CreateCatTable(vars = catvars, data = brca)
```
]
.pull-right[
```{r 08-Summaries-14, eval=T, echo = F, ref.label="tab1"}
```
]

---

.pull-left[
```{r tab2, eval = F, echo = T}
CreateContTable(vars = ctsvars, data = brca)
```
]
.pull-right[
```{r 08-Summaries-15, eval=T, echo = F, ref.label="tab2"}
```
]

---

.pull-left[
```{r tab2a, eval = F, echo = T}
brca <- brca %>% 
  rename(
    'Age'='Age.at.Initial.Pathologic.Diagnosis',
    'Last.Contact' = 'Days.to.Date.of.Last.Contact',
    'Death' = 'Days.to.date.of.Death'
  )
ctsvars <- brca %>% select_if(is.numeric) %>% names()
CreateContTable(vars = ctsvars, data = brca)

```
]
.pull-right[
```{r 08-Summaries-16, eval=T, echo = F, ref.label="tab2a"}
```
]

---

## Putting it together

.pull-left[
```{r tab3, eval = F, echo = T}
CreateTableOne(vars = c(catvars, ctsvars),
               data = brca)
```
]
.pull-right[
```{r 08-Summaries-17, eval=T, echo = F, ref.label="tab3"}
```
]

---

## Putting it together

.pull-left[
```{r tab4, eval = F, echo = T}
CreateTableOne(data = brca[,-1])
```
]
.pull-right[
```{r 08-Summaries-18, eval=T, echo = F, ref.label="tab4"}
```
]


---

class: middle, center

# Grouped summaries

---

.pull-left[
```{r grp1, eval = F, echo = T, warning=F, message=F}
brca %>% 
  group_by(ER.Status) %>% 
  summarize_at(vars(starts_with('NP')),
               mean)
```

There are missing values now, so we have to use `na.rm=T`.
]
.pull-right[
```{r 08-Summaries-19, eval=T, echo = F, ref.label="grp1", warning=F, message=F}
```
]

---

.pull-left[
```{r grp1a, eval = F, echo = T}
brca %>% 
  group_by(ER.Status) %>% 
  summarize_at(vars(starts_with('NP')),
               mean, na.rm=T) #<<
```

We still have a row for the missing values of ER.Status
]
.pull-right[
```{r 08-Summaries-20, eval=T, echo = F, ref.label="grp1a", warning=F, message=F}
```
]

---

.pull-left[
```{r grp1b, eval = F, echo = T}
brca %>% 
  filter(!is.na(ER.Status)) %>% #<<
  group_by(ER.Status) %>% 
  summarize_at(vars(starts_with('NP')),
               mean, na.rm=T)
```

How about reversing the rows and columns for readability

]
.pull-right[
```{r 08-Summaries-21, eval=T, echo = F, ref.label="grp1b"}
```
]

---

.pull-left[
```{r grp1c, eval = F, echo = T}
brca %>% 
  filter(!is.na(ER.Status)) %>% #<<
  group_by(ER.Status) %>% 
  summarize_at(vars(starts_with('NP')),
               mean, na.rm=T) %>% 
  tidyr::gather(ID, value, -ER.Status) %>% #<<
  spread(ER.Status, value) #<<
```
]
.pull-right[
```{r 08-Summaries-22, eval=T, echo = F, ref.label="grp1c"}
```
]

---

### Using `tableone`

```{r grp1d, eval = F, echo = T}
CreateTableOne(
  data = brca %>% filter(!is.na(ER.Status)),
  vars = brca %>% 
    select(starts_with('NP')) %>% 
    names(),
  strata = 'ER.Status',
  test = F)
```
```{r 08-Summaries-23, eval=T, echo = F, ref.label="grp1d"}
```

---
class: middle, center

# Comparing two groups

---

## The t-test

The t-test compares whether the mean of a variable differs between two groups. 

It does assume the normal distribution for the data, but is robust to deviations from normality

Do **not** test for normality before doing the t-test. It isn't necessary and screws up your error rates

---

## The t-test

In R, there is a convenient function `t.test`

```{r 08-Summaries-24}
t.test(NP_958782 ~ ER.Status, data = brca)
```

Read the code as 

"Do a t-test to see if (the mean of) `NP_958782` differs by `ER.Status`, where
both are taken from the data set `brca`"

You can read the `~` as "by", as in "t-test of NP_958782 by ER.Status"

---

## The t-test

The packge `broom` provides a function `tidy` that makes the results of these
statistical tests tidy.

```{r 08-Summaries-25}
t.test(NP_958782 ~ ER.Status, data=brca) %>% 
  broom::tidy()
```

--

```{r 08-Summaries-26, echo=F}
t.test(NP_958782 ~ ER.Status, data=brca)
```

---

## Using `broom`

The fact that `broom::tidy` makes the results of tests into tibbles is in fact
extremely useful in high-throughput work

.pull-left[
```{r test1, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status)
```
]
.pull-right[
```{r 08-Summaries-27, eval=T, echo = F, ref.label="test1"}
```
]

---

## Using `broom`

The fact that `broom::tidy` makes the results of tests into tibbles is in fact
extremely useful in high-throughput work

.pull-left[
```{r test2, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status) %>% 
  group_split(protein)
```
]
.pull-right[
```{r 08-Summaries-28, eval=T, echo = F, ref.label="test2"}
```
]

---

## Using `broom`

The fact that `broom::tidy` makes the results of tests into tibbles is in fact
extremely useful in high-throughput work

.pull-left[
```{r test2a, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status) %>% 
  split(.$protein) %>% #<<
  map(~broom::tidy(t.test(expression ~ ER.Status, 
                          data=.)))
```
]
.pull-right[
```{r 08-Summaries-29, eval=T, echo = F, ref.label="test2a"}
```
]

---

## Using `broom`

The fact that `broom::tidy` makes the results of tests into tibbles is in fact
extremely useful in high-throughput work

.pull-left[
```{r test2b, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status) %>% 
  split(.$protein) %>% #<<
  map(~broom::tidy(t.test(expression ~ ER.Status, 
                          data=.))) %>% 
  bind_rows(.id='Protein')
```
]
.pull-right[
```{r 08-Summaries-30, eval=T, echo = F, ref.label="test2b"}
```
]

---

## Using `broom`

The fact that `broom::tidy` makes the results of tests into tibbles is in fact
extremely useful in high-throughput work

.pull-left[
```{r test2c, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status) %>% 
  split(.$protein) %>% #<<
  map(~broom::tidy(t.test(expression ~ ER.Status, 
                          data=.))) %>% 
  bind_rows(.id='Protein') %>% 
  select(Protein, estimate, p.value, conf.low, conf.high)
```
]
.pull-right[
```{r 08-Summaries-31, eval=T, echo = F, ref.label="test2c"}
```
]

---
class: center, middle

# Back to testing

---

## Wilcoxon test, nonparametric t-test

```{r 08-Summaries-32}
wilcox.test(NP_958782 ~ ER.Status, data=brca) %>% 
  broom::tidy()
```

--

```{r 08-Summaries-33, echo=F}
wilcox.test(NP_958782 ~ ER.Status, data=brca)
```

---

## Wilcoxon test

.pull-left[
```{r test3, eval = F, echo = T}
brca %>% 
  select(ER.Status, starts_with('NP')) %>% 
  tidyr::gather(protein,expression, -ER.Status) %>% 
  split(.$protein) %>% 
  map(~broom::tidy(wilcox.test(expression ~ ER.Status, 
                          data=.))) %>% 
  bind_rows(.id='Protein') %>% 
  select(Protein, p.value)
```
]
.pull-right[
```{r 08-Summaries-34, eval=T, echo = F, ref.label="test3"}
```
]

---

## Using `tableone`


```{r table4, eval = F, echo = T}
CreateTableOne(
  data = brca %>% filter(!is.na(ER.Status)),
  vars = brca %>% 
    select(starts_with('NP')) %>% 
    names(),
  strata = 'ER.Status',
  test = T,
  testNormal = t.test
)
```
```{r 08-Summaries-35, eval=T, echo = F, ref.label="table4"}
```

--

This is not quite the same results as before

---

## Using `tableone`


```{r table4a, eval = F, echo = T}
CreateTableOne(
  data = brca %>% filter(!is.na(ER.Status)),
  vars = brca %>% 
    select(starts_with('NP')) %>% 
    names(),
  strata = 'ER.Status',
  test = T,
  testNormal = t.test,
  argsNormal = list(var.equal=F) #<<
)
```
```{r 08-Summaries-36, eval=T, echo = F, ref.label="table4a"}
```

---

## Tests for discrete data

Testing whether the distribution of a categorical variable differs by levels of 
another categorical variable can be done using either the Chi-square test (`chisq.test`) or the Fisher's test (`fisher.test`). Both require you to create a 2x2 table first.

```{r 08-Summaries-37}
fisher.test(table(brca$Tumor, brca$ER.Status))
```

---

## Tests for discrete data

Testing whether the distribution of a categorical variable differs by levels of 
another categorical variable can be done using either the Chi-square test (`chisq.test`) or the Fisher's test (`fisher.test`). Both require you to create a 2x2 table first.

```{r 08-Summaries-38}
chisq.test(table(brca$Tumor, brca$ER.Status))
```

---

## Tests for discrete data

We can use `broom::tidy` for either of these

```{r 08-Summaries-39}
chisq.test(table(brca$Tumor, brca$ER.Status)) %>% 
  broom::tidy()
```

---

## Using `tableone`

```{r 08-Summaries-40}
CreateCatTable(vars = c('Tumor','Node','Metastasis'),
               data = filter(brca, !is.na(ER.Status)),
               strata = 'ER.Status',
               test = T) # chisq.test
```

---

## Using `tableone`

```{r 08-Summaries-41}
c1 <- CreateCatTable(vars = c('Tumor','Node','Metastasis'),
               data = filter(brca, !is.na(ER.Status)),
               strata = 'ER.Status',
               test = T)
print(c1, exact = c('Tumor','Node','Metastasis')) # fisher.test
```
