---
title: "Displaying analytic results"
author: Abhijit Dasgupta
date: BIOF 339
---

```{r setup, include=F, child=here::here('slides/templates/setup.Rmd')}
```

```{r setup1, include=F}
library(ggpubr)
library(survival)
```


```{r, echo=FALSE, results='asis'}
update_header()
```

---
class: middle,center,inverse

# Comparing groups

---

```{r, echo=FALSE, results='asis'}
update_header('The `ggpubr` package')
```

---

The **ggpubr** package, which extends **ggplot2** functionality, has several 
functions that allow the computation and visualization of different statistical analyses

Under the hood, it's just fancy application of R for statistical tests and then 
translating the results to ggplot geometries.

---


.pull-left[
```{r boxes, echo=T, eval=F}
plt <- ggboxplot(penguins,
          x = 'species',
          y = 'body_mass_g',
          color = 'species',
          add='jitter',
          add.params = list(color='grey', size=0.5))
summary.stats <- penguins %>% 
  select(body_mass_g, species) %>% 
  group_by(species) %>% 
  get_summary_stats(type='common')
plt2 <- ggsummarytable(
  summary.stats, x = 'species', y = c('n','median','iqr')
) + 
    theme_minimal()+
    theme(panel.grid=element_blank(),
          axis.text.x = element_blank())+
    labs(x='', y='')
ggarrange(plt, plt2, ncol=1, heights=c(3,1))

```
]
.pull-right[
```{r, eval=T,echo=F, ref.label='boxes'}
```

]

---

.pull-left[
```{r box1, echo=T, eval=F}
(plt <- ggplot(penguins,
              aes(x=species,
                  y=body_mass_g,
                  color=species))+
  geom_boxplot()+
  geom_jitter(color='grey', size=0.5)+
  theme_classic()+
  theme(legend.position = 'top'))
```
]
.pull-right[
```{r, eval=T, echo=F, ref.label='box1'}

```

]

---

.pull-left[
```{r, ref.label='box1', echo=T, eval=F}
(plt <- ggplot(penguins,
              aes(x=species,
                  y=body_mass_g,
                  color=species))+
  geom_boxplot()+
  geom_jitter(color='grey')+
  theme_classic()+
  theme(legend.position = 'top'))
```

```{r box2, echo=T, eval=F}
summary.stats <- penguins %>% 
  select(body_mass_g, species) %>% 
  group_by(species) %>% 
  get_summary_stats(type='common')
(plt2 <- ggsummarytable(
  summary.stats, x = 'species', 
  y = c('n','median','iqr'),
  color='species'
  ) + 
    theme_minimal()+
    theme(panel.grid=element_blank(),
          axis.text.x = element_blank())+
    labs(x='', y=''))
```

]
.pull-right[
```{r, eval=T, echo=F, ref.label='box2'}

```

]

---

.pull-left[
```{r, ref.label='box1', echo=T, eval=F}

```
```{r, ref.label='box2',echo=T, eval=F}

```
```{r box3, eval=F, echo=T}
ggarrange(plt, plt2, ncol=1, heights = c(4,1))
```


]
.pull-right[
```{r, eval=T, echo=F, ref.label='box3'}

```

]

---

.pull-left[
```{r box4, eval = F, echo = T}
(summ_plt <- ggsummarystats(
  penguins,
  x = 'species',
  y = 'body_mass_g',
  ggfunc = ggboxplot, add='jitter',
  color='species',
  add.params=list(color='grey', size=0.5)
))
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="box4"}
```
]

---

.pull-left[
```{r, eval=F, echo=T, ref.label='box4'}

```

```{r box5, eval = F, echo = T}
my_comparisons <- list(c('Adelie','Chinstrap'),
                       c('Chinstrap','Gentoo'),
                       c('Adelie','Gentoo'))
summ_plt$main.plot <- summ_plt$main.plot + 
  stat_compare_means(comparisons = my_comparisons)+
  stat_compare_means(label.y = 8000)
summ_plt
```

-----
You can also add facets using the argument `facet.by` in `ggsummarystats`
]
.pull-right[
```{r, eval=T, echo = F, ref.label="box5"}
```
]

---

```{r corr, include=FALSE}

ggscatter(penguins, 
          x = 'bill_length_mm', 
          y = 'body_mass_g', 
          color='grey', 
          add='reg.line',
          add.params=list(color='blue')) +
  labs(x = 'Bill length (mm)',
       y = 'Body mass (g)') +
  stat_cor( label.x=30, label.y=6000, 
            label.sep='\n') + 
  facet_wrap(~species)
```

`r chunk_reveal('corr', title="Scatter plots")`

---

### Add tables to a graphic

.pull-left[
```{r dens1, eval = F, echo = T}
library(ggsci)
dens_plt <- ggplot(penguins, 
                   aes(x = body_mass_g))+
  geom_density(aes(fill=species))+
  scale_fill_jco(alpha=0.3)+
  theme_classic() +
  theme(axis.text.y = element_blank(),
        legend.position='top') 

stable <- desc_statby(penguins, 
                   measure.var='body_mass_g',
                   grps='species')
stable <- stable[,c('species','length','mean','sd')]
stable_plt <- ggtexttable(stable, rows=NULL)

ggarrange(dens_plt, stable_plt,
          ncol=1, heights=c(2,1))
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="dens1"}
```
]

---

### Add tables to a graphic

.pull-left[
```{r dens2, eval = F, echo = T}
library(ggsci)
dens_plt <- ggplot(penguins, 
                   aes(x = body_mass_g))+
  geom_density(aes(fill=species))+
  scale_fill_jco(alpha=0.3)+
  theme_classic() +
  theme(axis.text.y = element_blank(),
        legend.position='top') +
  stat_central_tendency( #<<
    aes(color=species), #<<
    type='mean', #<<
    geom='line', linetype=2 #<<
  ) + #<<
  scale_color_jco() #<<

stable <- desc_statby(penguins, 
                   measure.var='body_mass_g',
                   grps='species')
stable <- stable[,c('species','length','mean','sd')]
stable_plt <- ggtexttable(stable, rows=NULL)

ggarrange(dens_plt, stable_plt,
          ncol=1, heights=c(2,1))
```
]

---

```{r, echo=FALSE, results='asis'}
update_header('Modeling results')
```

---

Let's start with a somewhat naive multiple linear regression model

```{r}
m1 <- lm(body_mass_g ~ ., data=penguins)

```

.pull-right[
```{r, eval=T, echo = F, ref.label="dens2"}
```
]

---

.pull-left[
```{r ellipse1, eval = F, echo = T}
ggscatter(penguins,
          x='bill_length_mm',
          y='body_mass_g',
          color='species',
          shape='species',
          size=0.5,
          ellipse=TRUE)+
  stat_mean(aes(color=species, shape=species), size=4)
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="ellipse1"}
```
]

---

```{r, echo=FALSE, results='asis'}
update_header('## Model results')
```

---

Let's start with a fairly naive multiple regression model

```{r, echo=T}
m1 <- lm(body_mass_g ~., data=penguins)
summary(m1)$coef
```

---

```{r mod1, include=F}
library(gtsummary)
theme_gtsummary_compact()
tbl_regression(m1,
  label = list(bill_length_mm ~ 'Bill length (mm)',
               bill_depth_mm ~ 'Bill depth (mm)')) %>% 
  add_global_p() %>% 
  bold_p(t = 0.05) %>% 
  bold_labels() %>% 
  italicize_levels()
```

`r chunk_reveal('mod1')`

---

### Putting together multiple models

.pull-left[
```{r m1, echo=T, eval=F}
(gt_r1 <- glm(response ~ trt + grade, trial, 
              family = binomial) %>%
  tbl_regression(exponentiate = TRUE))
```
```{r m2, echo=F, eval=F}
(gt_r2 <- coxph(Surv(ttdeath, death) ~ trt + grade,
                trial) %>%
  tbl_regression(exponentiate = TRUE))
```
```{r m3, echo=F, eval=F}
(gt_t1 <- trial[c("trt", "grade")] %>% 
  tbl_summary(missing = "no") %>% 
  add_n() %>%
  modify_header(stat_0 ~ "**n (%)**") %>%
  modify_footnote(stat_0 ~ NA_character_))
```
```{r m4, echo=F, eval=F}
tbl_merge(
  list(gt_t1, gt_r1, gt_r2),
  tab_spanner = c(NA_character_, "**Tumor Response**", "**Time to Death**")
)
```
]
.pull-right[
```{r, eval=T, echo=F, ref.label='m1'}

```
]

---

### Putting together multiple models

.pull-left[
```{r, eval=F, echo=T, ref.label='m2'}
```

]
.pull-right[
```{r, eval=T, echo=F, ref.label='m2'}

```

]

---

### Putting together multiple models

.pull-left[
```{r, eval=F, echo=T, ref.label='m3'}

```

]
.pull-right[
```{r, eval=T, echo=F, ref.label='m3'}

```

]

---

### Putting together multiple models

```{r, eval=T, echo=T, ref.label='m4'}

```

---

### The finalfit package

```{r, echo=T}
explanatory <- setdiff(names(penguins), 'body_mass_g')
dependent <- 'body_mass_g'
t2 <- penguins %>% 
  finalfit::finalfit(dependent, explanatory, metrics=TRUE)
knitr::kable(t2[[1]], row.names = F)
```

---

### Other packages

+ **stargazer**
+ **sjPlot**

#### Utility packages

+ **gt**
+ **flextable**
* **huxtable**

---

```{r, echo=FALSE, results='asis'}
update_header('## Model results\n### Plotting results')
```

---

.pull-left[
```{r model1, echo=T, eval=F}
out <- broom::tidy(m1) %>% 
  slice(-1) %>% 
  mutate(lcb = estimate - 2*std.error,
         ucb = estimate + 2*std.error) %>% 
  select(term, estimate, lcb, ucb)
knitr::kable(out, digits = 2) %>% 
  kable_styling()

```
```{r model2, echo=F, eval=F}
(plt1 <- out %>% 
  ggplot(aes(x = term, y = estimate,
             ymin = lcb, ymax = ucb))+
  geom_pointrange()+
  geom_hline(yintercept=0, linetype=2)+
  xlab('') + geom_vline(xintercept=0, linetype=2)+
  coord_flip() + theme_classic())
```


]
.pull-right[
```{r, echo=F, eval=T, ref.label='model1'}

```

]

---

.pull-left[
```{r, ref.label='model1', echo=F, eval=F}

```
```{r , ref.label='model2', echo=T, eval=F}
```

]
.pull-right[
```{r, echo=F, eval=T, ref.label='model2', fig.height=5}

```

]

---

.pull-left[
```{r, echo=F, eval=F, ref.label='model1'}

```
```{r, echo=T, eval=F, ref.label='model2'}

```
```{r model3, echo=T, eval=F}
out <- out %>% 
  mutate(across(-term, round, 2)) %>% 
  mutate(ci = glue::glue('{estimate} ({lcb},{ucb})'))
plt2 <- ggplot(out, aes(x = term, y = 0))+
  geom_text(aes(label=ci), size=3, hjust=0)+
  coord_flip()+theme_void() + scale_y_continuous(limits = c(-0.1, 0.2))
ggpubr::ggarrange(plt1, plt2, nrow=1, widths=c(2,1), align='h')
```



]
.pull-right[
```{r, echo=F, eval=T, ref.label='model3', fig.height=5}

```

]

---

```{r, echo=FALSE, results='asis'}
update_header('## Model results \n### Survival analysis')
```

---

.pull-left[
```{r surv, echo=T, eval=F}
library(survival)
library(survminer)
pbc <- pbc %>% mutate(trt = as.factor(trt), stage = as.factor(stage))
m <- survfit(Surv(time, status==2)~trt, data=pbc)
ggsurvplot(m, pval=TRUE, risk.table=T)
```
]
.pull-right[
```{r, ref.label='surv', eval=T, echo=F}

```

]

---

.pull-left[
```{r, ref.label='surv', echo=T, eval=F}

```
```{r surv2, echo=T, eval=F}
m2 <- coxph(Surv(time, status==2) ~ trt, data = pbc)
gtsummary::tbl_regression(m2, exponentiate=T)
```
]
.pull-right[
```{r, ref.label='surv2', echo=F, eval=T}

```

]

---

.pull-left[
```{r, echo=T, eval=F}
m3 <- coxph(Surv(time, status==2) ~ trt + sex + 
              age + stage, data=pbc)
out <- broom::tidy(m3) %>% 
  mutate(lcb = estimate - 2*std.error,
         ucb = estimate + 2*std.error) %>% 
  mutate(across(c(estimate,lcb,ucb), exp))
gtsummary::tbl_regression(m3, exponentiate = T)
```
]
.pull-right[
```{r, echo=T}
out %>% 
  ggplot(aes(x = term, y = estimate,
             ymin = lcb, ymax=ucb))+
  geom_pointrange()+
  geom_hline(yintercept=1, linetype=2)+
  coord_flip()+
  theme_classic()
```

]
