---
title: Homework 1, Part 2
author: _________ ___________
date: "BIOF 339"
---


```{r, echo = FALSE, eval=TRUE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```


## Descriptive statistics

We'll start with the `airquality` data set that is in-built in R. 

1. The average temperature in June was `r mean(airquality$Temp[airquality$Month==6], na.rm=TRUE)`.
2. Solar radiation data is missing on `r sum(is.na(airquality[,"Solar.R"]))` days, or in `r 100 * sum(is.na(airquality[,"Solar.R"]))/nrow(airquality)` percent of all the days collected. 

We can also visualize the missing data patterns in this data set.

```{r, echo = TRUE, eval=TRUE}
library(pacman)
p_load('naniar') # This is a package for missing data
vis_miss(airquality) # see the documentation for vis_miss
```

Let's grab a more interesting data set. We will download and use the [Palmer Station penguins data set](doi:10.1371/journal.pone.0090081), which is in the form 
of an R package on GitHub. 

```{r, echo = TRUE, eval=TRUE}
library(pacman)
p_load('visdat') # Install and load visdat
p_install_gh('allisonhorst/palmerpenguins') 
p_load('palmerpenguins')
vis_dat(penguins)
```

