---
title: "Homework 5: Plotting"
author: "Solution"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(message=F, warning=F)
```


## Reading

1. Please read [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html#introduction-1) 
of R4DS. This is a really nice reference. If you're interested, look at [section 3.7](https://r4ds.had.co.nz/data-visualisation.html#statistical-transformations) for a deeper understanding of how different statistical transformations can be accommodated within different geometries in ggplot2. 

## Exercises

<ol>
<li> Let's look at the  [breast cancer data set](data/clinical_data_breast_cancer_hw.csv) we'be been using.  

<ol>
  <li> Create a bar plot of the frequencies of ER status (Hint: `geom_bar`. How would you modify the dataset
  so that there is no separate bar for NA's?
```{r}
library(tidyverse)
library(rio)
brca <- import('data/clinical_data_breast_cancer_hw.csv')
brca <- brca %>% filter(ER.Status != 'Indeterminate')
ggplot(brca, aes(x = ER.Status))+geom_bar()
```

  <li> Color each bar by the frequencies of HER2 status within each group. (Hint: you can add `aes(color='HER2.Status')` or `aes(fill='HER2.Status')` to the `geom_bar` statement. What is the difference in these two choices?)
```{r}
ggplot(brca, aes(x = ER.Status)) + geom_bar(aes(fill=HER2.Final.Status))
```
  
  <li> I would like to see a graph showing the relative frequencies of HER2 status within each ER group. Read the documentation `?geom_bar` as well as section 3.8 of R4DS to find this solution. 
```{r}

ggplot(brca, aes(x = ER.Status)) + geom_bar(aes(fill=HER2.Final.Status), position='fill')
```
  
</ol>

<li> The `diamonds` dataset is included in the `ggplot2` package. 
<ol>
<li> Add a column to the diamonds dataset that is the volume of each diamond (multiplying the `x`, `y` and `z` variables)
```{r}
mydiamonds <- diamonds %>% 
  mutate(vol = x*y*z)
```

<li> Plot a scatter plot of diamond price against diamond volume
```{r}
ggplot(mydiamonds, aes(x =vol, y = price)) + geom_point()
```

<li> Modify the plot to add colors based on the cut of each diamond
```{r}
ggplot(mydiamonds, aes(x =vol, y = price, color=cut)) + geom_point()
```

<li> Modify the previous plot to construct a facetted plot where each facet is based on a color. 
```{r}
ggplot(mydiamonds, aes(x = vol, y = price))+
  geom_point()+
  facet_wrap(~color)
```

</ol>
</ol>

### Extra credit

The [breast cancer expression](data/BreastCancer_Expression.xlsx) dataset contains 
log-expression levels of 10 probes for the same individuals that are in the clinical dataset.

1. Modify the dataset appropriately and draw a facetted plot where each facet is a histogram of the expression levels of each probe. 
```{r}
brca_expression <- import('data/BreastCancer_Expression.xlsx')
brca_expression %>% gather(probe, expression, starts_with('NP')) %>% 
  ggplot(aes(x = expression))+geom_histogram()+
  facet_wrap(~probe)

```


