---
title: "Homework 2"
author: "BIOF 339"
output: html_document 

---

```{r setup, eval = FALSE, include=FALSE} 
# First modify this chunk to activate the packages you will need for this HW.
# Ensure you have `tidyverse`, `knitr` and `rmarkdown` installed
# Add statements like `library(tidyverse)`

library(_____)
p_load(_______)
p_load(_______)

# add more p_load statements 
# if you need to. Or you can do it in one go with 
# 
# p_load(char = c("______", "________", "________"))
```

# Instructions

You are expected to edit this R Markdown document, especially the R code chunks. Once
you're done, please change `eval=F` to `eval=T`  in each chunk in order to make the code run. Most of the 
R chunks are fill-in-the-blanks, with a few left empty for you to completely fill in. 

Your submission for this week will comprise 2 files:

1. This file, after editing to complete the assignments
1. The HTML file obtained by knitting this file

# Exercises

1. Fix the code in the following code to ensure they work without errors.

```{r, eval = F}
my_variable <-  1:5
my_var1able
```

```{r, eval = F}
mean(airquality$wind)
```

```{r, eval = F}
table(iris$Sepal_Length)
```


2. Download the data "clinical_data_breast_cancer_modified.csv" from the data link on Canvas and save it
in the `data` folder of your RStudio Project for this class.

  a. Import the data into R, calling it `breast_cancer`. 

```{r, eval = F}
# Code to import data file here

__________ <- __::______("data/clinical_data_breast_cancer_modified.csv")

```
  
  b. Check the data types for each of the variables
```{r, eval = F}
# Check for data types

____(breast_cancer)

```
  c. Convert the Gender, ER Status, PR Status and HER2 status variables to factors
```{r, eval = F}
# Convert here

breast_cancer$Gender <- as.factor(breast_cancer$Gender)

## You can repeat this for the other variables, making sure you spell them properly. Or you can comment 
## proceed as follows:

breast_cancer <- breast_cancer %>%
    mutate(Gender = as.factor(Gender), 
           ______= ________(_______),
           ..... # fill this in with the other variables
          )

## Comment out one of the two strategies for your submission

```
  d. Identify any other variables that might be the wrong type and convert them
```{r}
# Convert any new variables here
```
  e. Verify that all the variables in `breast_cancer` match the modifications you made
```{r, echo = F, eval = F}
# Don't edit this chunk. It is for verification.
breast_cancer %>% 
  dplyr::summarize_all(class) %>% 
  tidyr::pivot_longer(names_to = "Variable", values_to = "Class", cols = everything()) %>% 
  knitr::kable()
```

3. The dataset "clinical_data_breast_cancer_hw.csv" provides clinical and
pathological data for 105 breast cancer patients. 

a. Read this data into R, calling it `brca_data`. 

```{r, eval=FALSE}
brca_data <- ____::________('data/clinical_data_breast_cancer_hw.csv')
```

b. This data is in tidy format, but needs cleaning. 
    a. Evaluate whether all the variables are of the correct type. If not, evaluate why, and correct the problem.

```{r, eval=FALSE}
___(brca_data)

## Add code here to correct any problems in the data set


```

b. For the molecular results for ER, PR and HER2, any entry that is not Positive or Negative should be changed to a missing value.

```{r, eval=FALSE}
## There are two approaches here. The first looks at the data and finds out the actual values that aren't Positive or Negative
brca_data1 <- brca_data %>%
    mutate(ER.Status = recode_factor(ER.Status, Indeterminate = NA_character_), 
    mutate(________ = recode_factor(_______,  '_________' = NA_character_))

## There is an error in the above code, and a couple of solutions.
  
## The other is to run the same function on the columns for ER, PR and HER2 status

clean_markers <- function(x){
  x <- recode_factor(x, Positive = 'Positive', _______ = '________', .default = NA_character_)
  return(x)
 }

 brca_data2 <- brca_data %>%
    mutate(across(c(____, ____, ____), ________))

## Verify that both data sets are identical

all(brca_data1 == brca_data2, na.rm=TRUE)

```


c. Create a new dataset that contains patient ID, ER, PR and HER2 status. 

```{r, eval=FALSE}
brca_data3 <- _______(brca_data, ______, ______, _______, _______)
```


d. Creating new variables (what `dplyr` function will you use for all of these?): ___________

e. Create a variable giving the TNM status of each patient. The T, N and M statuses are given separately. I want a single variable encoded as, for example, "T2N0M0". [Hint: The function `paste` is your friend]

```{r, eval=FALSE}
brca_data <- brca_data %>%
    ______(tnm_status = ___________(Tumor, Node, Metastasis, sep = ""))
brca_data3$tnm_status <- brca_data$tnm_status
```

f. Create a variable that divides the patients into 3 categories based on molecular profiles [Hint: The function `case_when` might help]:
        - Luminal (ER positive and/or PR positive)
        - HER2 (HER2 positive)
        - Basal-like (ER, PR and HER2 negative)

```{r, eval=FALSE}

brca_data3 <- brca_data3 %>%
    mutate(mol_cat = case_when( # fill in the next 4 lines

      
      
      )
```


d. Create a variable that equals the time of death if they are dead (see the variable `Vital.Status`) or the time of last contact if they are alive. This is a common computation for survival analysis studies, and is called the overall survival time. [Hint: the function `ifelse` might be useful]

```{r, eval=FALSE}
## Fill this in yourself. Time to start leaving the nest
brca_data <- brca_data %>% 
  mutate(event_time = ___________(Vital.Status == 1, _____________, _______________))
brca_data3$event_time <- brca_data$event_time

```


Save the cleaned breast cancer dataset as `brca_cleaned`. You can save this to your computer using `saveRDS(brca_cleaned, file="<a filename of your choice>.rds")`. We'll be using this dataset again when we do plots and modeling.


    
