---
title: "Loops, Maps"
author: Abhijit Dasgupta
date: Fall, 2019

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '#> ')
library(tidyverse)
```

layout: true

<div class="my-header">
<span>BIOF339, Fall, 2019</span></div>

---

# Where we are

+ Got a start on plotting and creating panelled graphs with ggplot2
+ Can modify a data set somewhat
    - `dplyr` verbs (mutate, filter, select, separate, unite)
    - joins
    - gather/spread

---

# Repetitive copying

For HW 6 (sorry about the mess), you had to copy and paste multiple times to get
things done

- Had to do same processing on multiple data sets
- Had to do same graphs from multiple data sets

---
class: middle, center

# For loops and Maps

---

## For loops

For-loops are a computational structure that allows you to do the same thing
repeatedly over a loop with some index. 

The basic structure is

```{r 07-Maps-Table1-1, eval=F}
for (variable in vector) {
  <code to execute for each iteration>
}
```

---

## For loops

.pull-left[
Using numeric indices

```{r 07-Maps-Table1-2, eval = F, echo = T}
library(fs)
sites <- c('Brain','Colon','Esophagus','Lung','Oral')
dats <- list() # Initialize empty list
for(i in 1:length(sites)){
  dats[[i]] <- 
    read_csv(path('data',
                             paste0(sites[i], '.csv')), 
                        skip=4)
}
```
]
.pull-right[
Using names

```{r 07-Maps-Table1-3, eval=T, echo = T, message=F, warning=F}
library(fs)
sites <- c('Brain','Colon','Esophagus','Lung','Oral')
dats <- list() # Initialize empty list
for(n in sites){
  dats[[n]] <- 
    read_csv(path('data',
                  paste0(n, '.csv')), 
             skip=4)
}
```
]

---

## Lists

Directly using lists has efficiency advantages. `rio` can load all the datasets
into a list, for example.

```{r 07-Maps-Table1-4}
dats <- rio::import_list(path('data', paste0(sites,'.csv')))
names(dats)
str(dats[['Brain']])
```

--

Recall, lists are the most generic buckets in R. Elements of lists can be anything. 
To use `map` it's best that each element of the input list be of the same type
---

## Maps

`map` is like a for-loop, but strictly for lists. It is more efficient than for-loops. The basic template is:

```{r 07-Maps-Table1-5, eval=F}
map(<list>, <function>, <function arguments>)
```

For example, if we want to take out the first row of each dataset and make sure all 
the variables are numeric, we could do:

```{r 07-Maps-Table1-6, warning=F}
dats <- map(dats, function(d){
  d %>% slice(-1) %>%  # remove first row
    mutate_all( as.numeric)
})
str(dats[['Brain']])
```

---

## Maps

`map` is like a for-loop, but strictly for lists. It is more efficient than for-loops. The basic template is:

```{r 07-Maps-Table1-7, eval=F}
map(<list>, <function>, <function arguments>)
```

For example, if we want to take out the first row of each dataset and make sure all 
the variables are numeric, we could do:

```{r 07-Maps-Table1-8, warning=F, eval=F}
dats <- map(dats, function(d){
  d %>% slice(-1) %>%  # remove first row
    mutate_all( as.numeric)
})
str(dats[['Brain']])
```

The argument for the function inside the `map` function is an element of the list. In this case, it is a data frame.

The output of `map` is a list the same length as the input list.

---

## Maps

I don't like the names with spaces, so I can just apply a function to each data set to fix that.

```{r 07-Maps-Table1-9}
dats <- map(dats, janitor::clean_names)
str(dats[['Oral']])
```

---

## Maps

Now let's split up by sexes

```{r 07-Maps-Table1-10}
dats_all <- map(dats, select, year_of_diagnosis, ends_with('sexes'))
dats_male <- map(dats, select, year_of_diagnosis, ends_with('_males'))
dats_female <- map(dats, select, year_of_diagnosis, ends_with('females'))
str(dats_all[['Esophagus']])
```

--

Here I used the form `map(<list>, <function>, <function arguments>)`.

Earlier I had used `map(<list>,<function definition>)` and `map(<list>, <function>)` with no (i.e., default) arguments.

---

## Maps

Let's make the column headers of each dataset reflect the site, so that when we join we can keep the 
sites separate

```{r 07-Maps-Table1-11}
for(n in sites){
  names(dats_all[[n]]) <- str_replace(names(dats_all[[n]]), 'both_sexes',n)
  names(dats_male[[n]]) <- str_replace(names(dats_male[[n]]), 'male',n)
  names(dats_female[[n]]) <- str_replace(names(dats_female[[n]]), 'female',n)
}
names(dats_all[['Esophagus']])
```


---
## Higher order maps

When we joined these data sets, we had to repeatedly use `left_join` to create the final data set. There is a shortcut to this repeated operation of a function with two inputs as applied to a list successively.

```{r 07-Maps-Table1-12, message=F}
dats2_all <- Reduce(left_join, dats_all)
dats2_male <- Reduce(left_join, dats_male)
dats2_female <- Reduce(left_join, dats_female)
```

--
> Could we have used a for loop or map here? Sure, but it makes it harder to read IMO.

--
```{r 07-Maps-Table1-13}
str(dats2_all)
```

---

## Maps

Next, we want to separate the races from the sites, after a `gather`. The `all_races` will pose a problem if we split on `_`. Let's fix that.

```{r 07-Maps-Table1-14}
names(dats2_all) <- str_replace(names(dats2_all), 'all_races','allraces')
names(dats2_male) <- str_replace(names(dats2_male), 'all_races','allraces')
names(dats2_female) <- str_replace(names(dats2_female), 'all_races','allraces')
```

---

## Maps

Now, for each of these , we need to gather then separate. We'll put the data sets in a list first

```{r 07-Maps-Table1-15}
dats2 <- list('both'=dats2_all, 'male'=dats2_male, 'female'=dats2_female)
dats2 <- map(dats2,
             function(d){
               d %>% tidyr::gather(variable, rate, -year_of_diagnosis) %>% 
                 separate(variable, c('race','site'), sep='_')
             })
str(dats2[['both']])
```

---

## Final graphing

Now we're in a position to do the graphing. 

.pull-left[
```{r plt1, eval = F, echo = T}
pltlist <- dats2[['both']] %>% group_split(race) %>% 
  map(function(d) {ggplot(d, 
                          aes(x = year_of_diagnosis,
                              y = rate,
                              color=site))+
  geom_point(show.legend = F) })
cowplot::plot_grid(plotlist=pltlist, ncol=1, 
                   labels=c('Both','Males','Females'))
```

I'm using quite advanced R here, but hopefully you'll learn by example.

`group_split` splits the dataset by the values of the grouping variable into a list

(Yes, your homework asked for a different panel placement)
]
.pull-right[
```{r 07-Maps-Table1-16, eval=T, echo = F, ref.label="plt1"}
```
]
```{r 07-Maps-Table1-17}

```

