---
title: "Data validation and exploration"
author: Abhijit Dasgupta
date: Fall, 2019

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '#> ')
library(tidyverse)
```

layout: true

<div class="my-header">
<span>BIOF339, Fall, 2019</span></div>

---

## Plan today

- Dynamic exploration of data
- Data validation
- Missing data evaluation

---
class: middle, center

# Why go back to this?

---

## This is important!!

+ Most of the time in an analysis is spent understanding and cleaning data
+ Recognize that  unless you've ended up with good-quality data, the rest of the analyses are moot
+ This is tedious, careful, non-sexy work
    - Hard to tell your boss you're still fixing the data
    - No real results yet
    - But essential to understanding the appropriate analyses and the tweaks you may need.
    
---

## What does a dataset look like?

.pull-left[
```{r v1, eval = F, echo = T}
library(tidyverse)
library(visdat)
beaches <- rio::import('data/sydneybeaches3.csv')
vis_dat(beaches)
```
]
.pull-right[
```{r, eval = T, echo = F, ref.label = "v1"}
```
]

---

## What does a dataset look like?

.pull-left[
```{r v2, eval = F, echo = T}
brca <- rio::import('data/clinical_data_breast_cancer_modified.csv')
vis_dat(brca)
```
]
.pull-right[
```{r, eval = T, echo = F, ref.label = "v2"}
```
]

---

## What does a dataset look like?

.pull-left[
```{r v3, eval = F, echo = T}
vis_dat(airquality)
```

These plots give a nice insight into 

1. data types
1. Missing data patterns (more on this later)

]
.pull-right[
```{r, eval = T, echo = F, ref.label = "v3"}
```
]

---
class: middle, center

# Let's get a bit more quantitative

---

## `summary` and `str`/`glimpse` are a first pass

.pull-left[
```{r}
summary(airquality)
```
]
.pull-right[
```{r}
glimpse(airquality)
```

]

---

## Validating data values

+ We can certainly be reactive by just describing the data and looking for anomalies. 
+ For larger data or multiple data files it makes sense to be proactive and catch errors that you want to avoid, before exploring for new errors. 
+ The `assertthat` package provides nice tools to do this

--

> **Note to self:** I don't do this enough. This is a good defensive programming technique that can catch crucial problems that aren't always automatically flagged as errors

---

## Being assertive

```{r, error=T}
library(assertthat)
assert_that(all(between(airquality$Day, 1, 31) ))
assert_that(is.factor(mpg$manufacturer))
assert_that(all(beaches$season_name %in% c('Summer','Winter','Spring', 'Fall')))

```

---

## Being assertive

+ `assert_that` generates an error, which will stop things
+ `see_if` does the same validation, but just generates a `TRUE/FALSE`, which you can capture

```{r}
see_if(is.factor(mpg$manufacturer))
```

+ `validate_that` generates `TRUE` if the assertion is true, otherwise generates a string explaining the error

```{r}
validate_that(is.factor(mpg$manufacturer))
validate_that(is.character(mpg$manufacturer))
```

---

## Being assertive

You can even write your own validation functions and custom messages

```{r, error=TRUE}
is_odd <- function(x){
    assert_that(is.numeric(x), length(x)==1)
    x %% 2 == 1
}
assert_that(is_odd(2))

on_failure(is_odd) <- function(call, env) {
  paste0(deparse(call$x), " is even") # This is a R trick
}
assert_that(is_odd(2))

is_odd(1:2)
```

---
class: middle,center

# Missing data

---

## Missing data

R denotes missing data as `NA`, and supplies several functions to deal with missing data.

The most fundamental is `is.na`, which gives a TRUE/FALSE answer

```{r}
is.na(NA)
is.na(25)
```

---

## Missing data

When we get a new dataset, it's useful to get a sense of the missingness

```{r}
mpg %>% summarize_all(function(x) sum(is.na(x)))
```

---

## Missing data

The `naniar` package allows a tidyverse-compatible way to deal with missing data

```{r}
library(naniar)
weather <- rio::import('data/weather.csv')
all_complete(mpg)
all_complete(weather)
weather %>% summarize_all(pct_complete)
```

---


## Missing data

```{r}
gg_miss_case(weather, show_pct = T)
```

---

## Missing data

```{r}
gg_miss_var(weather, show_pct=T)
```

## Are there patterns to the missing data

+ Most analyses assume that data is either 
    - Missing completely at random (MCAR)
    - Missing at random (MAR)
+ MCAR means
    - The missing data is just a random subset of the data
+ MAR means
    - Whether data is missing is related to values of some other variable(s)
    - If we control for those variable(s), the missing data would form a random subset of each of those data subsets defined by unique values of these variables

---

## Are there patterns to the missing data

#### MAR or MCAR allows us to ignore the missing data, since it doesn't bias our analyses
#### If data are **not** MCAR or MAR, we really need to understand the issing data mechanism and how that might affect our results. 
---

## Co-patterns of missingness

.pull-left[
```{r v4, eval = T, echo = T}
gg_miss_upset(airquality)
```
]
.pull-right[
```{r v5, eval = T, echo = T}
gg_miss_upset(riskfactors)
```
]

---

## Co-patterns of missingness

.pull-left[
```{r d1, fig.height=3, message=T}
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
geom_point()
```
]
.pull-right[
```{r, fig.height=3}
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
 geom_miss_point()
```
]

---

## Co-patterns of missingness

```{r, warning=F}
gg_miss_fct(x = riskfactors, fct = marital)
```

---

## Replacing missing data

`tidyr` has  a function `replace_na` which will replace all missing values with some particular value. 

In the weather dataset, values are missing generally because there wasn't recorded rainfall on a day. So these values should really be 0

```{r}
weather1 <- weather %>% mutate(d1 = replace_na(d1, 0))
pct_miss(weather1$d1)
```

---

### Question: How would you replace all the missing values with 0?

--

```{r, results = 'hide'}
weather %>% mutate_all(function(x) replace_na(x, 0))
```

--

### How would you replace the missing values with the mean of the variable?

--

```{r, results = 'hide'}
weather %>% mutate_if(is.numeric, function(x) replace_na(x, mean(x, na.rm=T)))
```

---
